Fig3: Correlations in gene expression

### 2020-04-25 analysis/figure-2.Rmd
now with fig 4

source("../R/themes.R")

@@

## Candidate Genes (from literature and relevant GO terms)

```{r GO}
source("../R/wrangledata.R")
```

@@

## variance stabilized gene expression  (vsd)

```{r vsds}

vsd_path <- "../results/DEseq2/"   # path to the data
vsd_files <- dir(vsd_path, pattern = "*vsd.csv") # get file names
vsd_pathfiles <- paste0(vsd_path, vsd_files)
vsd_files

allvsd <- vsd_pathfiles %>%
  setNames(nm = .) %>% 
  map_df(~read_csv(.x), .id = "file_name")  %>% 
  dplyr::rename("gene" = "X1") %>% 
  pivot_longer(cols = L.G118_female_gonad_control:y98.o50.x_male_pituitary_inc.d3, 
               names_to = "samples", values_to = "counts")

getcandidatevsd <- function(whichgenes, whichtissue, whichsex){
  candidates  <- allvsd %>%
    filter(gene %in% whichgenes) %>%
    dplyr::mutate(sextissue = sapply(strsplit(file_name, '_vsd.csv'), "[", 1)) %>%
    dplyr::mutate(sextissue = sapply(strsplit(sextissue, '../results/DEseq2/'), "[", 2)) %>%
    dplyr::mutate(sex = sapply(strsplit(sextissue, '\\_'), "[", 1),
                tissue = sapply(strsplit(sextissue, '\\_'), "[", 2),
                treatment = sapply(strsplit(samples, '\\_'), "[", 4)) %>%
    dplyr::mutate(treatment = sapply(strsplit(treatment, '.NYNO'), "[", 1)) %>%
    dplyr::select(sex, tissue, treatment, gene, samples, counts) %>%
    filter(tissue == whichtissue, sex %in% whichsex)  %>%
    drop_na()
  candidates$treatment <- factor(candidates$treatment, levels = alllevels)
  return(candidates)
}

candidategenes <- GOgenesLong %>% distinct(gene) %>% pull(gene)

hypvsd <- getcandidatevsd(candidategenes, "hypothalamus", sexlevels)
pitvsd <- getcandidatevsd(candidategenes, "pituitary", sexlevels)
gonvsd <- getcandidatevsd(candidategenes, "gonad", sexlevels)
candidatevsd <- rbind(hypvsd, pitvsd)
candidatevsd <- rbind(candidatevsd, gonvsd)
head(candidatevsd)
```

@@

>>>>

source("../R/themes.R")
source("../R/wrangledata.R")

@@

@@

## Candidate variance stabilized gene expression (vsds)

@@

# `candidatevsd` loaded with source(../R/wrangleata.R)

### 2020-04-25 analysis/figure-3.Rmd
fig tweeks
now with fig 4

---
title: "Fig3"
author: "Rayna M Harris"
date: "3/26/2020"
output: md_document
---

# all things box plots

## candidate genes

```{r}
library(tidyverse)
library(ggtext)
library(cowplot)
library(ggpubr)

source("../R/themes.R")

knitr::opts_chunk$set(echo = TRUE, message = F, fig.path = "../figures/")

candidategenes <- c("OXT", "AVP", "GNRH1", "GNRHR", "CGNRH-R",
                    "AR", "POMC", "AGRP", 
                       "CRH", "AVPR1A", "AVPR1B", "AVPR2","VIP",
                       "CYP19A1", "DRD1", "DRD2", "PRL", "PRLR", "SOX9", 
                    "ESR1","ESR2", "LBH", "CDK1", "BRCA1",
                    "PTEN", "CREBBP", "FOS", "JUN", "EGR1",
                     "BDNF", "GRM2","GRIA1",
                    "KCNJ5", "CISH", "PTGER3", "CEBPD", "ZBTB16", 
                    "DIO3", "DIO2", "DIO1",

                     "VIPR1", "VIPR2", "NPY", "LH", "FSH", "FSHB",
                    "JAK2", "HCRT", "TRH", "TSHB",
                    "MC3R", "MC4R", "MC5R",  "NR3C1", "NR3C2",
                    "STAT5B","NPVF")

candidategenesslim <- c("OXT", "AVP", "GNRH1", "GNRHR", 
                    "AR",  "CYP19A1", 
                    "AVPR1A", "AVPR1B", "AVPR2","VIP",
                    "DRD1", "DRD2", 
                    "PRL", "PRLR",  
                    "ESR1","ESR2", "LBH",  
                    "FOS", "JUN", "EGR1", "BDNF")

# candidategenesslim: AR, AVP, AVPR1A, AVPR1B, AVPR2, BDNF, CYP19A1, DRD1, EGR1, ESR1, ESR2, FOS, GNRH1, GNRHR, JUN, LBH, OXT, PRL, PRLR, VIP
# DEG candidategenesslim: AR, AVPR1A, AVPR1B, AVPR2, BDNF, CYP19A1, DRD1, ESR1, FOS, GNRHR, LBH, PRL, PRLR, 
# non-DEG candidategenesslim: AVP, AVPR1B, ESR2,  GNRH1, JUN, OXT, VIP

hormonepapergenes <- c("GNRH1", "CGNRH-R", "GNRHR",  
                       "FSHB", "FSHR", "LHCGR", # no LSH
                       "VIP", "VIPR1",
                       "ESR1", "AR", "CYP19A1", "HSD17B1", # receptors synthesis
                       "PGR" , "HSD3B2", 
                       "PRL", "PRLR")
```

## variance stabilized gene expression  (vsd)

```{r }
geneids <- read_csv("../metadata/00_geneinfo.csv") %>% select(-X1)

vsd_path <- "../results/DEseq2/"   # path to the data
vsd_files <- dir(vsd_path, pattern = "*vsd.csv") # get file names
vsd_pathfiles <- paste0(vsd_path, vsd_files)
vsd_files

allvsd <- vsd_pathfiles %>%
  setNames(nm = .) %>% 
  map_df(~read_csv(.x), .id = "file_name")  %>% 
  dplyr::rename("gene" = "X1") %>% 
  pivot_longer(cols = L.G118_female_gonad_control:y98.o50.x_male_pituitary_inc.d3, 
               names_to = "samples", values_to = "counts")

getcandidatevsd <- function(whichgenes, whichtissue, whichsex){
  candidates  <- allvsd %>%
    filter(gene %in% whichgenes) %>%
    dplyr::mutate(sextissue = sapply(strsplit(file_name, '_vsd.csv'), "[", 1)) %>%
    dplyr::mutate(sextissue = sapply(strsplit(sextissue, '../results/DEseq2/'), "[", 2)) %>%
    dplyr::mutate(sex = sapply(strsplit(sextissue, '\\_'), "[", 1),
                tissue = sapply(strsplit(sextissue, '\\_'), "[", 2),
                treatment = sapply(strsplit(samples, '\\_'), "[", 4)) %>%
    dplyr::mutate(treatment = sapply(strsplit(treatment, '.NYNO'), "[", 1)) %>%
    dplyr::select(sex, tissue, treatment, gene, samples, counts) %>%
    filter(tissue == whichtissue, sex %in% whichsex)  %>%
    drop_na()
  candidates$treatment <- factor(candidates$treatment, levels = alllevels)
  return(candidates)
}

hypvsd <- getcandidatevsd(hormonepapergenes, "hypothalamus", sexlevels)
pitvsd <- getcandidatevsd(hormonepapergenes, "pituitary", sexlevels)
gonvsd <- getcandidatevsd(hormonepapergenes, "gonad", sexlevels)
head(hypvsd)
tail(gonvsd)

candidatevsd <- rbind(hypvsd, pitvsd)
candidatevsd <- rbind(candidatevsd, gonvsd)

unique(candidatevsd$gene)
```

## Figs

```{r fig3, fig.width = 7}
# summary DEG results from DESeq2
candidateDEGS <- read_csv("../../musicalgenes/data/allDEG.csv") %>%
  select(-X1) %>%
  filter(gene %in% hormonepapergenes)
head(candidateDEGS)

# get genes to plot
candidateDEGS %>% filter(tissue == "hypothalamus", sex == "female", comparison != "control_bldg") 
candidateDEGS %>% filter(tissue == "hypothalamus", sex == "male", comparison != "control_bldg") 
candidateDEGS %>% filter(tissue == "pituitary", sex == "male", comparison != "control_bldg") 
candidateDEGS %>% filter(tissue == "pituitary", sex == "female", comparison != "control_bldg") 
candidateDEGS %>% filter(tissue == "gonad", sex == "female", comparison != "control_bldg")

# summarize all differences
candidateDEGtable <- candidateDEGS %>%
  mutate(posneg = ifelse(lfc >= 0, "+", "-")) %>%
  mutate(geneposneg = paste(gene, posneg, sep = "")) %>%
  select(sex, tissue, comparison, geneposneg) %>%
  group_by(sex, tissue, comparison) %>%
   arrange(geneposneg) %>%
  summarize(genes = str_c(geneposneg, collapse = " ")) %>% 
  pivot_wider(names_from = comparison, values_from = genes)  %>% 
  ungroup(tissue) %>%
  mutate(tissue = factor(tissue, levels =  tissuelevel))  %>% 
  arrange(sex, tissue)  %>% 
  select(sex, tissue, bldg_lay, lay_inc.d3, inc.d9_inc.d17, hatch_n5, control_bldg)
candidateDEGtable

newboxplot <- function(df, mygenes, whichsex, whichstage){

  p <- df %>%
    filter(gene %in% mygenes, 
           sex %in% whichsex,
           treatment %in% whichstage) %>%
    droplevels() %>%
    mutate(tissue = factor(tissue, levels = tissuelevel)) %>%
    ggplot(aes(x = treatment, y = counts, fill = treatment, color = sex)) +
    geom_boxplot() +
    geom_jitter(size = 0.5, width = 0.1) +
    facet_grid(tissue~gene, scales = "free_y") +
    theme_B3() +
    scale_color_manual(values = allcolors) +
    scale_fill_manual(values = allcolors) +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 45, hjust = 1),
          strip.text.x = element_text(face = "italic"),
          axis.title.x = element_blank())  +
    labs(x = "Parental stage",
         y = "variance stabilized expression")
  return(p)
}

a1 <- newboxplot(hypvsd, c("PGR",  "CYP19A1", "LHCGR" ), "female", c( "hatch", "n5")) + 
  theme(strip.text.y = element_blank()) + labs(subtitle = "Females")
a2 <- newboxplot(hypvsd, c("AR"), "male", c( "hatch", "n5")) + 
  theme(axis.title.y = element_blank()) + labs(subtitle = "Males")

b1 <- newboxplot(gonvsd, c("HSD3B2"), "female", c( "bldg", "lay"))  +
  theme(strip.text.y = element_blank(),
        axis.title.y = element_blank()) + labs(subtitle = "Females")
b2 <- newboxplot(gonvsd, c("VIPR1",  "PRLR" ), "female", c( "lay", "inc.d3"))  + 
  theme(axis.title.y = element_blank()) + labs(subtitle = " ")

ab <- plot_grid(a1,a2,b1,b2, rel_widths = c(6,2.5,2,4), labels = c("a", " ", "b", " "), label_size = 8, nrow = 1)

c1 <- newboxplot(pitvsd, c( "ESR1", "GNRHR"), "female", c("bldg", "lay",  "inc.d3")) 
c2 <- newboxplot(pitvsd, c( "PRL"), "female", c("inc.d9", "inc.d17", "hatch", "n5")) 
c3 <- newboxplot(pitvsd, c("VIPR1",  "PRL" ), "male", c( "inc.d9", "inc.d17"))

c <- plot_grid(c1 + theme(strip.text.y = element_blank()) + labs(subtitle = "Females"),
          c2 + theme(strip.text.y = element_blank(),
                    axis.title.y = element_blank()) + labs(subtitle = " ") ,
          c3 + theme(axis.title.y = element_blank()) + labs(subtitle = "Males"), 
          nrow = 1, rel_widths = c(6,4,4),
          labels = c("c"), label_size = 8)

plot_grid(ab,c, nrow = 2)

# correlations
library(corrr)

plotcorrs <- function(whichtissue){

  x <- candidatevsd %>%
    filter(tissue == whichtissue)  %>%
    pivot_wider(names_from = gene, values_from = counts) %>%
    select(-sex, - tissue, - treatment, -samples) %>%
    correlate()
  print(fashion(x))

  p <- rplot(x, colors = c("#0571b0", "#92c5de", "#f7f7f7","#f7f7f7","#f7f7f7", "#f4a582", "#ca0020")) +
    theme_B3() +
    theme(legend.position = "bottom",
          axis.text.x = element_text(angle = 45, hjust = 1),
          axis.text = element_text(face = "italic")) +
    labs(subtitle = whichtissue)
  return(p)
}

a <- plotcorrs("hypothalamus") 
b <- plotcorrs("pituitary") + theme(axis.text.y = element_blank())
c <- plotcorrs("gonad") + theme(axis.text.y = element_blank())

plot_grid(a,b,c, nrow = 1, rel_widths = c(1.2,1,1))

@@

```{r writefiles}
#write.csv(candidatevsd, "../../musicalgenes/data/candidatecounts.csv")
#write.csv(candidatevsd, "../results/candidatecounts.csv")
write.csv(candidateDEGtable, "../results/candidateDEGtable.csv")

>>>>

---
title: "Fig3: Correlations in gene expression"
author: "Rayna M Harris"
date: "3/26/2020"
output: md_document
---

# Correlations in gene expression

```{r setup}
library(tidyverse)
library(cowplot)
library(corrr)
library(ggpubr)

source("../R/themes.R")
source("../R/functions.R")
source("../R/wrangledata.R")

knitr::opts_chunk$set(echo = TRUE, message = F, fig.path = "../figures/")
```

## data wrangle

```{r data}

# load `candidatevsd` with `source("../R/wrangledata.R")`
head(candidatevsd)

curleychampagnegenes <- curleychampagnegenes %>% pull(gene)

# widen for correlations

candidatevsdwide <- candidatevsd  %>%
    pivot_wider(names_from = gene, values_from = counts) 
head(candidatevsdwide)

makecorrdf <- function(whichsex, whichtissue, whichgenes){

  corrrdf <- candidatevsd %>%
    filter(sex == whichsex, tissue == whichtissue,
           gene %in% whichgenes) %>%
    pivot_wider(names_from = gene, values_from = counts) %>%
    select(-sex,-tissue, -treatment, -samples) %>%
    correlate() %>%
    rearrange()
  print(head(corrrdf))
  return(corrrdf)
}

c1 <- makecorrdf("female", "hypothalamus", curleychampagnegenes)  
c2 <- makecorrdf("female", "pituitary", curleychampagnegenes)  
c3 <- makecorrdf("female", "gonad", curleychampagnegenes)

c4 <- makecorrdf("male", "hypothalamus", curleychampagnegenes)  
c5 <- makecorrdf("male", "pituitary", curleychampagnegenes)  
c6 <- makecorrdf("male", "gonad", curleychampagnegenes)

subsetcandidatevsdwide <- function(whichsex, whichtissue){
  df <- candidatevsdwide %>%
    filter(sex == whichsex, tissue == whichtissue,
           treatment != "control") 
  return(df)
}

FH <- subsetcandidatevsdwide("female", "hypothalamus")
FP <- subsetcandidatevsdwide("female", "pituitary")
FG <- subsetcandidatevsdwide("female", "gonad")
MH <- subsetcandidatevsdwide("male", "hypothalamus")
MP <- subsetcandidatevsdwide("male", "pituitary")
MG <- subsetcandidatevsdwide("male", "gonad") 
```

### figure

```{r fig3, fig.width = 7.25, fig.height = 7.25}

plotcorrplot <- function(df, subtitle){  
  p <- df %>%
    rplot2() +
    theme_B3() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1),
          axis.text = element_text(face = "italic"))

  return(p)
}

p1 <- plotcorrplot(c1, "FH") + theme(legend.position = "none", axis.text.x = element_blank()) + labs(subtitle = "Hypothalamus")
p2 <- plotcorrplot(c2, "FP") + theme(legend.position = "none", axis.text.x = element_blank()) + 
  labs(subtitle = "Pituitary") 
p3 <- plotcorrplot(c3, "FG") + theme(legend.position = "none") + labs(subtitle = "Gonad")

p4 <- plotcorrplot(c4, "MH") + theme(legend.position = "none", axis.text.y = element_blank(), axis.text.x = element_blank()) + labs(subtitle = " ")
p5 <- plotcorrplot(c5, "MP") + theme(legend.position = "none", axis.text.y = element_blank(), axis.text.x = element_blank())  +
  labs(subtitle = " ") 
p6 <- plotcorrplot(c6, " ") + theme(legend.position = "none", axis.text.y = element_blank()) + labs(subtitle = " ")

a <- plot_grid(p1 + labs(title = "Female"),
               p4 + labs(title = "Male"),
               p2, p5, p3, p6, nrow = 3, rel_widths = c(1.2,1),
               rel_heights = c(1,1,1.2))

scattercorrelations <- function(df, gene1, gene2, mylinecolor, myxlab, myylab){
  p <- ggplot(df, aes(x = gene1, y = gene2)) +
    labs(subtitle = " ", x = myxlab, y = myylab) + 
    scale_color_manual(values = allcolors) +
    geom_smooth(method = "lm",  color = mylinecolor) +
    geom_point(aes(color = treatment))  +
    theme_B3() +  
    theme(axis.title = element_text(face = "italic"), 
            legend.position = "none") +
    stat_cor( size = 2)
  return(p)
}

# "female" = "#969696", "male" = "#525252"

p7 <- scattercorrelations(FH, FH$DRD1, FH$HTR2C, "#969696", "DRD1", "HTR2C") + labs(subtitle = "Hypothalamus")
p8 <- scattercorrelations(FP, FP$AVPR1A, FP$CRHR1, "#969696", "AVPR1A", "CRHR1")  + labs(subtitle = "Pituitary")
p9 <- scattercorrelations(FG, FG$ESR2, FG$CRHBP, "#969696", "ESR2", "CRHBP")   + labs(subtitle = "Gonad")

p10 <- scattercorrelations(MH, MH$DRD1, MH$HTR2C, "#525252", "DRD1", "HTR2C") + theme(axis.title.y = element_blank())
p11 <- scattercorrelations(MP, MP$AVPR1A, MP$CRHR1, "#525252", "AVPR1A", "CRHR1") + theme(axis.title.y = element_blank())
p12 <- scattercorrelations(MG, MG$ESR2, MG$CRHBP, "#525252", "ESR2", "CRHBP")  + theme(axis.title.y = element_blank())

b <- plot_grid(p7 + labs(title = "Female"), 
               p10 + labs(title = "Male"), 
               p8, p11, p9, p12, nrow = 3, rel_heights = c(1,1,1.2))

ab <- plot_grid(a,b, rel_widths = c(1.5,1), labels = c("A", "B"), label_size = 8)

p13 <- plotcorrplot(c1, "FH") + theme(legend.position = "bottom")
p14 <- scattercorrelations(FH, FH$DRD1, FH$HTR2C, "#969696", "DRD1", "HTR2C") + 
  theme(legend.position = "bottom",
        legend.title = element_blank())

legend1 <- p13 %>% get_legend()
legend2 <- p14 %>% get_legend()

legends <- plot_grid(legend1, legend2, rel_widths = c(1.5,1))

plot_grid(ab, legends, nrow = 2, rel_heights = c(1,0.1))

@@
